---
title: "PS270 Final Project"
author: "Akrama Mohamed"
description: "My final project"

---


## Project thoughts

I am interested in exploring data related to making predictions or collecting data. 

## Final Project Milestone 2:

How may achieving a higher education influence voter turnout in elections?

For my research, I would like to investigate how achieving a higher education would mean a greater voter turnout. ( How may achieving a higher education influence voter turnout in elections?) This hypothesis is rooted in the theory that individuals who may have earned their college degrees or higher than compared to individuals for example who are high-school graduates; have a greater awareness of the election associated with higher voter turnout. Additionally, higher levels of education typically provide individuals with a better understanding of political systems, the importance of civic participation, and policies. With this exposure to diverse sources of information, individuals may feel empowered or influenced to cast their votes, than those individuals who may not have gotten the opportunity for a higher education. In this research, the primary explanatory variable is education level, which is the calculation of people's achieved education measured in the number of years of schooling. Variation in this variable is crucial, and the data set includes a wide range of education levels, allowing us to capture differences across educational attainment groups.Furthermore, the outcome variable is the voter turnout, with 1 being an indication of individual voting and 0 for did not vote, a two-category measure determining whether an individual voted in the most recent election. To conclude, regarding observed patterns, it can be noted that if there is a consistent link between higher levels of education and increased voter turnout, it would provide evidence in support of the proposed hypothesis. Conversely, if there is no correlation between education level and voter turnout, it would refute the hypothesis, suggesting that other variables may account for voting behaviors.


## Final Project Milestone 3:

```{r}
library(ggplot2)
library(tidyverse)
library(readr)
library(dplyr)
turnout <- read_csv("index_files/turnouts.csv") 
```

```{r}
getwd()
setwd("/Users/ikramohamed/Desktop")
getwd
```

```{r}
ggplot(turnout, aes(x = Year, y = Outcome, color = Education)) +
  geom_line() + 
  geom_point() +
  labs(
    title = "Voter Turnout by Education Level in the United States (1964–2016)", 
    caption = "Sources : College Board; US Census Bureau 
    © Statista 2024 Some college or associate degree", 
    x = "Years",
    y = "Percentage of Voter Turnout (%)",
    color = "Education Level"
  ) +
  theme_minimal()
turnout
```
This is a plot showcasing the voting rates among citizens during presidential elections in the United States from 1964 to 2016, by education level, which does show the achieving a higher education does in fact influence the voter turnout in the United States. In other words those who have obtained a bachelor degree or higher have a higher voting turnout rates than compared to any other education level, according to the tables percentages and plot outcome. 

## Final Project Milestone 4:
```{r}
library(dplyr)
library(knitr)

turnout_table <- turnout |> 
  group_by(Education) |>
  summarize(
    Mean_Percentage = mean(Outcome),
    SD_Percentage = sd(Outcome),
    Min_Percentage = min(Outcome),
    Max_Percentage = max(Outcome)
  )

knitr::kable(turnout_table, 
            digits = 2, 
            col.names = c("Education Level", "Mean", "SD", "Min", "Max"),
            caption = "Summary of Voter Turnout by Education Level (1964-2016)")
```

My research question focused on examining how may achieving a higher education influence voter turnout in elections? The data presented in the table and the accompanying graph demonstrate that individuals with higher education, such as a bachelor's degree, have a higher average voter turnout compared to those with lower education levels. The table shows that individuals who have completed a bachelor's degree or have some college experience have the highest average voter turnout rates, approximately 81% and 69%, respectively. In contrast, individuals who have only a high school diploma or did not graduate have much lower averages, at 59% and 45%. Furthermore, those who have completed some college also exhibit the highest maximum voter turnout, reaching 88% and 82%. Overall, these findings support the claim that individuals who have attained higher education are more likely to participate in voting.


## Final Report

Introduction:

For my research, I aim to investigate the relationship between higher education and voter turnout. My hypothesis is: How does achieving a higher education influence voter turnout in elections? This hypothesis is based on the theory that individuals with college degrees or higher tend to have greater awareness of elections compared to individuals with lower education experience, leading to increased voter turnout. This is important because it will help us understand just how much education level affects voter turnout over time.

Data Section:

This research data is provided from a 2024 statistic published by the Statista Research Department shows the voting rates among citizens during presidential elections in the United States from 1964 to 2016, categorized by education level. The voting rates serve as the dependent variable, while education level is categorized as the independent variable. The data further proves our hypothesis with the indication that citizens with a bachelor degree or higher had increasingly higher voting percentages than those with lower education levels. This research uses a cross-sectional design, focusing on voter turnout and education levels at a specific point in time. It does not track changes in these variables at the individual level over time. Instead, it examines how different education levels correlate with voter turnout across the years.


Plot:
```{r}
turnout_summary <- turnout |>
  group_by(Year) |>
  summarize(Average_Percentage = mean(Outcome, na.rm = TRUE))

ggplot(turnout_summary, aes(x = Year, y = Average_Percentage)) +
  geom_line() + 
  geom_point() +
  labs(
    title = "Average Voter Turnout in the United States (1964–2016)", 
    x = "Years",
    y = "Average Percentage of Voter Turnout (%)",
    caption = "Sources: College Board; US Census Bureau"
  ) +
  theme_minimal()
```


Result Section:

```{r}
turnout_model <- lm(Outcome ~ Education, data = turnout)
turnout_model
summary(turnout_model)

library(broom)
model_results <- tidy(turnout_model)
knitr::kable(model_results, 
             digits = 3,
             caption = "Regression of Voter Turnout by Year and Education Level")
```

```{r}
ggplot(turnout, aes(x = Education, y = Outcome, fill = Education)) +
  stat_summary(fun = mean, geom = "bar") +
  labs(
    title = "Average Voter Turnout by Education Level",
    subtitle = "United States (1964–2016)",
    x = "Education Level",
    y = "Voter Turnout (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    plot.margin = margin(b = 40, l = 10, r = 10, t = 10)
  )
```

Both the regression ran with our data and the plot results illustrates the trends in voter turnout percentages from 1964 to 2016 for individuals categorized by education levels. Those with higher education (Bachelor's Degree or higher) consistently exhibit the highest voter turnout, while individuals with lower education levels (e.g., Grade 9 to 11) show the lowest turnout. Over time, turnout for all groups has experienced a slight decline, with the steepest reductions observed among individuals with less education.

Furthermore, the regression results support the insights derived from the plot, indicating significant differences in voter turnout based on education levels. The intercept of 80.950 represents the estimated voter turnout for the reference category—individuals with a Bachelor's Degree or higher—in a baseline year, serving as the benchmark for comparison. Specifically, individuals with Grade 9 to 11 education have a voter turnout that is 35.486 percentage points lower than the reference group, with a t-statistic of -13.917 and a p-value less than 0.001, indicating statistical significance. Those with a High School Diploma experience a turnout that is 21.614 percentage points lower than the reference group, also highly significant, with a t-statistic of -8.477 and a p-value less than 0.001. Additionally, individuals with some college education or an Associate Degree show a turnout that is 11.543 percentage points lower than those with a Bachelor's Degree or higher, which is statistically significant (t = -4.527, p < 0.001).

The statistical significance of all coefficients is evident from the p-values (all < 0.05), strongly supporting the hypothesis that individuals with higher education are more likely to vote. However, it is important to note that the regression design does not allow for causal inference, as factors such as the political climate, regional differences, and socio-economic status could confound the observed relationships. Without additional controls or a randomized design, we cannot claim that education causes higher voter turnout. Taking into fact that this data showed an outcome of a majority constant standard error results this does have its down side and problems for it could be a violation of assumptions, misleading robutness of estimates, and altogether not reflect the true variability of the data.The analysis has several limitations, including missing data or unmeasured confounders that may introduce bias, such as regional differences in voter laws or socio-economic factors. Additionally, the model assumes a linear relationship between education and voter turnout, which may oversimplify the reality. To improve the analysis, incorporating additional covariates—such as age, income, or state-level policies—and considering longitudinal or experimental designs could strengthen causal inference. Expanding the data-set to include geographic identifiers would also allow for the creation of detailed spatial maps, enhancing the interpret-ability of trends.



Conclusion:
In conclusion, the analysis strongly supports the hypothesis that voter turnout in the United States is influenced by education level, with significant differences among educational groups. Individuals with higher education (Bachelor’s Degree or higher) show higher voter turnout and a slower decline over time compared to those with lower education levels, particularly those with a Grade 9 to 11 education, who experience a more significant decrease. The regression model highlights that the relationship between time and voter turnout varies by education level. However, the cross-sectional nature of the data limits causal conclusions, as external factors like political shifts and economic events could influence trends. Additionally, confounding variables such as race, age, and income were not included. A longitudinal study with detailed individual-level data could better address these limitations and provide stronger insights into the factors affecting voter turnout. Incorporating variables like political ideology and regional differences would further enhance the analysis.






